---
title: "Exploring Global Trends in Life Expectancy and GDP"
autor: "Neoh Yi Theng"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(repos = c(CRAN = "https://cran.rstudio.com/"))
```

# Introduction

This project analyzes the 'Gapminder' dataset on R to explore global trends in life expectancy and GDP.

# Data Preparation

### Load the libraries and data

```{r, echo=TRUE, message=FALSE, warning=FALSE}

install.packages("gapminder")
install.packages("dplyr")
install.packages("ggplot2")
install.packages("shiny")

library(gapminder)
library(dplyr)
library(ggplot2)
library(shiny)
library(ggcorrplot)
library(rmarkdown)

data <- gapminder
```

### Data Summary

```{r}
summary(data)
```

### Check for missing and duplicate values

```{r}
sum(is.na(data))

duplicates <- data %>% 
  group_by(country, year) %>% 
  summarise(n = n()) %>% 
  filter(n > 1)
nrow(duplicates)
```

### Data structure

```{r}
str(data)
```

### Year range

```{r}
year_range <- range(data$year)
year_range
```

# Data Analysis

### Wordlwide metrics by year

```{r}
data_stats <- data %>%
  group_by(year) %>%
  summarise(
    mean_lifeExp = mean(lifeExp),
    sd_lifeExp = sd(lifeExp),
    median_lifeExp = median(lifeExp),
    mean_pop = mean(pop),
    sd_pop = sd(pop),
    median_pop = median(pop),
    mean_gdpPercap = mean(gdpPercap),
    sd_gdpPercap = sd(gdpPercap),
    median_gdpPercap = median(gdpPercap)
  )
data_stats
```

### Quartiles for GDP per Capita and Life Expentancy

```{r}
quantiles <- data %>%
  summarise(
    q1_gdpPercap = quantile(gdpPercap, 0.25),
    q3_gdpPercap = quantile(gdpPercap, 0.75),
    q1_lifeExp = quantile(lifeExp, 0.25),
    q3_lifeExp = quantile(lifeExp, 0.75)
  )
quantiles
```

### Life expectancy changes across continents

```{r}
continent_lifeExp <- data %>%
  group_by(continent, year) %>%
  summarise(mean_lifeExp = mean(lifeExp))
continent_lifeExp
```

### GDP per Capita changes across continents

```{r}
continent_gdpPercap <- data %>%
  group_by(continent, year) %>%
  summarise(mean_gdpPercap = mean(gdpPercap))
continent_gdpPercap
```

### Highest Life Expectancy and GDP per Capita by year

```{r}
highest_lifeExp <- data %>%
  group_by(year) %>%
  filter(gdpPercap == max(gdpPercap)) %>%
  select(year, country, continent, gdpPercap, lifeExp) %>%
  arrange(year)

highest_gdpPercap <- data %>%
  group_by(year) %>%
  filter(lifeExp == max(lifeExp)) %>%
  select(year, country, continent, gdpPercap, lifeExp) %>%
  arrange(year)
```

### Correlation between GDP per Capita and Life Expectancy

```{r}
correlation <- cor(data$lifeExp, data$gdpPercap, method = 'pearson')
cat("Pearson correlation coefficient between life expectancy and GDP per Capita is", correlation)
```

# Statistical Testing

### ANOVA Test for GDP per Capita across continents

```{r}
anova_gdpPercap <- aov(gdpPercap ~ continent, data = data)
summary_result <- summary(anova_gdpPercap)
p_value <- summary_result[[1]]$`Pr(>F)`[1]

cat("The p-value for the ANOVA test is", p_value, "\n")

if (p_value < 0.05) {
  cat("The differences in GDP per capita across continents are statistically significant.")
} else {
  cat("The differences in GDP per capita across continents are not statistically significant.")
}
```

### T-test Comparing Asia & Europe GDP per Capita

```{r}
asia_gdp <- data %>% filter(continent == "Asia") %>% pull(gdpPercap)
europe_gdp <- data %>% filter(continent == "Europe") %>% pull(gdpPercap)

t_test_result <- t.test(asia_gdp, europe_gdp)
t_test_result

# Interpretation
p_value <- t_test_result$p.value
cat("The p-value for the t-test comparing GDP per capita between Asia and Europe is", p_value, "\n")

if (p_value < 0.05) {
  cat("The difference in GDP per capita between Asia and Europe is statistically significant.")
} else {
  cat("The difference in GDP per capita between Asia and Europe is not statistically significant.")
}
```

# Trend Analysis

### Life expectancy over time

```{r}
ggplot(data, aes(x = year, y = lifeExp, color = continent)) +
  geom_line(stat = "summary", fun = mean) +
  labs(title = "Life Expectancy Over Time by Continent", x = "Year", y = "Life Expectancy")
```

### GDP per Capita over time

```{r}
ggplot(data, aes(x = year, y = gdpPercap, color = continent)) +
  geom_line(stat = "summary", fun = mean) +
  labs(title = "GDP per Capita Over Time", x = "Year", y = "GDP per Capita")
```

### Correlation Matrix

```{r}
numeric_data <- data %>% select(lifeExp, gdpPercap, pop)
correlation_matrix <- cor(numeric_data)
print(correlation_matrix)

# Visualize Correlation Matrix
ggcorrplot(correlation_matrix, lab = TRUE)
```

# Data Visualizations

### Boxplot of GDP per Capita by Continent

```{r}
ggplot(data, aes(x = continent, y = gdpPercap, fill = continent)) +
  geom_boxplot() +
  labs(title = "GDP per Capita by Continent", y = "GDP per Capita", x = "Continent")
```

### Boxplot of Life Expectancy by Continent

```{r}
ggplot(data, aes(x = continent, y = lifeExp, fill = continent)) +
  geom_boxplot() +
  labs(title = "Life Expectancy by Continent", y = "Life Expectancy", x = "Continent")
```

### Scatterplot of GDP per Capita and Life Expectancy

```{r}
ggplot(data, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  labs(title = "Life Expectancy vs GDP per Capita", x = "GDP per Capita", y = "Life Expectancy")
```

# Filtering & removing outliers

```{r}
# Calculate IQR for GDP per Capita and Life Expectancy
gdp_iqr <- IQR(data$gdpPercap)
gdp_q1 <- quantile(data$gdpPercap, 0.25)
gdp_q3 <- quantile(data$gdpPercap, 0.75)

lifeExp_iqr <- IQR(data$lifeExp)
lifeExp_q1 <- quantile(data$lifeExp, 0.25)
lifeExp_q3 <- quantile(data$lifeExp, 0.75)

# Filter outliers
filtered_data <- data %>%
  filter(
    gdpPercap > (gdp_q1 - 1.5 * gdp_iqr) & gdpPercap < (gdp_q3 + 1.5 * gdp_iqr),
    lifeExp > (lifeExp_q1 - 1.5 * lifeExp_iqr) & lifeExp < (lifeExp_q3 + 1.5 * lifeExp_iqr)
  )
```

### Boxplot of GDP per Capita by Continent **without outliers**

```{r}
ggplot(filtered_data, aes(x = continent, y = gdpPercap, fill = continent)) +
  geom_boxplot() +
  labs(title = "GDP per Capita by Continent (Without Outliers)", y = "GDP per Capita", x = "Continent")
```

### Boxplot of Life Expectancy by Continent **without outliers**

```{r}
ggplot(filtered_data, aes(x = continent, y = lifeExp, fill = continent)) +
  geom_boxplot() +
  labs(title = "Life Expectancy by Continent (Without Outliers)", y = "Life Expectancy", x = "Continent")
```

### Scatterplot of GDP per Capita and Life Expectancy **without outliers**

```{r}
ggplot(filtered_data, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  labs(title = "Life Expectancy vs GDP per Capita (Without Outliers)", x = "GDP per Capita", y = "Life Expectancy")
```
